\vspace{-2cm} Inference on the Effects of Observed Features \\ in Latent Space Models for Networks

### 2017-03-10 response.tex

\section{Response Memo}\label{response-memo}

\subsection{Response to Reviewer 1}\label{response-to-reviewer-1}

\begin{quote}
``The paper does not accurately characterize or analytically understand
the latent space approaches employed by the work that is cited. For
example Dorff and Ward (2013) do not use LATENTNET. LATENTNET is
extremely outdated and is not the software used in much of the other
work cited herein. This undermines much of the criticisms in the papers
and raises a red flag that suggests the authors do not grasp the
modeling approaches as deeply as they would present.''
\end{quote}

While we appreciate that the \texttt{latentnet} package is not the
implementation most commonly used in political science, it is used
outside of political science. We have made this point more clearly in
the revised manuscript. Most importantly, our criticism is not specific
to the implementation in \texttt{latentnet}, and we argue (again, more
clearly in the revised manuscript), that our argument does not depend on
the specific software implementation or parameterization used, and hence
this criticism does not speak to our point, which is that that the LSM
cannot adjust for unobserved confounders, even in a synthetic scenario
designed to make this as easy as possible.

\begin{quote}
``The authors are not up-to-date on the state of latent modeling in the
field and have ignored more recent work (which addresses many of the
concerns raised by the authors) particularly by Peter Hoff and Shahryar
Minhas. Additionally, the progress made by the newly developed AMEN
software is completely overlooked by this article.''
\end{quote}

We have added discussion of the AMEN package, which we were aware of,
but again, does not apply to our argument for the same reason.

\begin{quote}
``In that vein, authors identify a problem (e.g., collinearity between
latent distance parameters and exogenous variables in the latentnet
framework) and do not solve it nor discuss any recent work that tries to
address this problem.''
\end{quote}

Our point is that when the measured covariates are \emph{not}
independent of the unobserved structure in the network, that estimating
that structure using latent variables cannot adjust for any confounding.
We are unaware of any work that attempts to tackle this problem and we
furthermore consider the problem as we've posed it intractable for
theoretical reasons, which we have elucidated more clearly in our
revised version. In generally latent variables cannot be used to adjust
the effects of measured covariates. In the case where the measured
covariates are independent of any unmeasured structure, the LSM does no
worse than standard models, as we show in our simulation.

\subsection{Response to Reviewer 2}\label{response-to-reviewer-2}

\begin{quote}
``Yet, in their efforts to contextualize their research the authors seem
to disregard or not be aware of notable differences between existing
latent variable approaches. Specifically, much of the work they cite
(e.g., Ward Siverson and Cao 2007, Ward, Ahlquist and Rozenas 2013,
etc.) does not even use the type of latent variable model that the
authors focus on in their paper. In fact, with the exception of a 2012
piece by Kirkland, I am not aware of any published political science
work that actually uses the latent space model for network analysis the
authors study here.''
\end{quote}

While this is true, it is orthogonal to our point, which is that the
LSM, regardless of how the latent variables are parameterized, cannot
adjust for unobserved network confounders. We chose the first
formulation of the LSM to make our synthetic demonstration as
understandable as possible. We argue that it shows that even in the most
ideal scenario, the LSM cannot adjust for confounders that are
unmeasured. We have made the point that this behavior is independent of
the parameterization of the latent variables more clear in the revised
manuscript. While we certainly could have chosen any number of the
variations of the LSM for our simulation, absent clear reasons to do so
we do not believe that it is reasonable to expend the substantial
computational resources necessary to expand the simulation.

\begin{quote}
``The difference between these approaches has been shown to be
consequential in a number of works. Hoff \& Ward (2004) and Hoff (2005)
detail the ways in which the GBME is different than the latent distance
model that is employed in the latentnet package. Hoff (2008) explicitly
discusses how the latent distance model confounds transitivity and
stochastic equivalence (two types of third order interdependencies) in
problematic ways that the latent factor model is able to avoid. The
issue this paper seems most concerned with showing is that the latent
distance model does not solve the problem of bias and inferential error
relative to GLM for observed covariates. They do so through a well
executed simulation exercise and the results are certainly compelling in
favor of showcasing a shortcoming of the latent distance model. At the
same time, this result is not very surprising. For the latent distance
model we expect there to be some collinearity between covariates for
which there is homophily and latent position. In theory, a weak prior on
the covariate coefficients and a strong (tight) prior on the
coefficients should minimize this effect, but at the cost of less
expressiveness for latent positions.''
\end{quote}

We agree that the parameterization of the latent variables can affect
what can be learned about omitted structure in the network, as
suggested. However, this is unrelated to our point which is that it is
not possible to adjust the estimates of the effect of observed
covariates using a latent variable model that can represent unobserved
structure. We argue more clearly in the revised manuscript that this is
not a property of the Euclidean distance formulation we chose for
expository simplicity.

### 2017-03-16 manuscript.tex
finished response to R1

Latent variable inference, generally conceived, presents the possibility of representing unmeasured data in statistical models. The LSM, introduced by \citet{hoff2002latent}, is used to estimate the effect of covariates in the presence of latent network structure. Here the distance function $|z_i - z_j|$ represents latent network structure as homophily with respect to latent variables.  The distance function is additively combined with a regression on observed dyadic covariates, $x_{ij}$, to form a linear predictor for tie prediction.  As with a GLM, a link function, $g^{-1}$, maps this linear predictor to the appropriate edge distribution.

$$\mathbb{E}(y_{ij} | x_{ij}) = g^{-1}(\alpha + \beta x_{ij} - |z_i - z_j|)$$

However, it is unclear that the introduction of a latent space decreases the expected error for the parameter(s) of interest when aspects of the network structure (e.g., homophily) that are unmeasured, are correlated with measured covariates. There are two reasons that using the LSM may lead to increased error. First, the latent configurations inferred may result in a representation of the network wherein a node's position in the latent space is spuriously correlated with the observed covariates. Inferring latent variables that are correlated with observed covariates would lead to reduced efficiency in estimating observed covariate effects. Second, if the unobserved (i.e., latent) network structure is truly correlated with the observed covariates, the unobserved structure that can be correlated with the observed variable may be attributed to the observed variable. Under this condition the latent space parameters are used to model sources of variation that cannot be attributed to the observed covariate, and inferences regarding the observed covariates remain subject to omitted variable bias.

@@

The LSM has seen a substantial amount of subsequent development and extension. The latent space has been represented as a $k$-dimensional Euclildean space and by latent factors \citep{hoff2002latent, hoff2009multiplicative}. Additional structure has been introduced by adding random effects, which, for example, may involve sender or receiver specific effects for directed networks which capture differential activity rates amongst nodes \citep{hoff2003random}. Within this framework \cite{westveld2011mixed} models dynamic network data by treating the latent space as a stochastic process. \cite{handcock2007model} enable the LSM to model clustering that is not representable as homophily (i.e., stochastic equivalence) by combining the LSM with latent cluster models. \cite{hoff2008modeling} shows that the LSM and latent cluster models are special cases of an ``eigenmodel.'' That is, an eigendecomposition of a symmetric sociomatrix can be used to represent both latent space and cluster models, but not vice-versa. Most of these developments have represented new forms for the latent space and/or the further development of the LSM as a Bayesian hierarchical model. In the next section we propose an approach to specifying the Bayesian prior in order to improve inference regarding covariate parameters.

>>>>

Latent variable inference, generally conceived, presents the possibility of representing unmeasured data in statistical models. The LSM, introduced by \citet{hoff2002latent}, is used to estimate the effect of covariates in the presence of latent network structure. Here the distance function $|z_i - z_j|$ represents latent network structure as homophily with respect to latent variables.  The distance function is additively combined with a regression on observed dyadic covariates, $x_{ij}$, to form a linear predictor for tie prediction.  As with a GLM, a link function, $g^{-1}$, maps this linear predictor to the appropriate edge distribution, giving

$$\mathbb{E}(y_{ij} | x_{ij}) = g^{-1}(\alpha + \beta x_{ij} - |z_i - z_j|).$$

This ``Euclidean'' LSM is the original form proposed by \citet{hoff2002latent}, and, as far as we can tell, the most commonly used specification in the literature. However, the LSM has been extended in several important directions. The  The latent space has been represented as a $k$-dimensional Euclildean space and by latent factors with a multiplicative metric \citep{hoff2002latent, hoff2004modeling, hoff_2005_jasa, hoff2009multiplicative}. Additional structure has been introduced by adding random effects, which, for example, may involve sender or receiver specific effects for directed networks which capture differential activity rates amongst nodes \citep{hoff2003random}. Within this framework \cite{westveld2011mixed} models dynamic network data by treating the latent space as a stochastic process. \cite{handcock2007model} enable the LSM to model clustering that is not representable as homophily (i.e., stochastic equivalence) by combining the LSM with latent cluster models. \cite{hoff2008modeling} shows that the LSM and latent cluster models are special cases of an ``eigenmodel.'' That is, an eigendecomposition of a symmetric sociomatrix can be used to represent both latent space and cluster models, but not vice-versa. \citet{hoff2015dyadic} consolidates, extends, and implements several of these extensions in the \texttt{amen} package \citep{amen} for \R.

Despite the power of the LSM in representing network structure that is not modeled by the observed covariates, and its performance in predictive ties predicting ties, we have reason to doubt the LSM's capacity to adjust for confounding network structure. There are two reasons that using the LSM may lead to increased error---one may result in Type 1 inferential error, and other Type 2. First, the latent configurations inferred may result in a representation of the network wherein a node's position in the latent space is spuriously correlated with the observed covariates. Inferring latent variables that are correlated with observed covariates would lead to reduced efficiency in estimating observed covariate effects, and result in a high Type 2 error rate.  Second, if the unobserved (i.e., latent) network structure is truly correlated with the observed covariates, the unobserved structure that can be correlated with the observed variable may be attributed to the observed variable. Under this condition the latent space parameters are used to model sources of variation that cannot be attributed to the observed covariate, and inferences regarding the observed covariates remain subject to omitted variable bias, which leads to an inflated Type 1 inferential error rate.

Recent work on the LSM has explicitly addressed the potential problem of Type 2 error. The AMEN framework for latent variable modeling of networks is explicitly designed to model structure in the residuals---structure that is leftover after accounting for the observed covariates. \citet[p. 43]{hoff2015dyadic} notes that the latent factors represent patterns in the network that, ``aren't explained by the known regressors.'' \citet[pp. 12--13]{minhas2016inferential} also describe how AMEN is designed such that the multiplicative effects, ``capture higher-order dependence patterns that are left over [in the stochastic linear predictor] after accounting for any known covariate information.''  The approach incorporated in AMEN is effective at avoiding correlation between the observed covariates and latent factors, which would result in Type 2 error. However, there have been no methodological innovations to avoid Type 1 error in latent variable modeling with networks. Below we review several studies in which researchers have relied on the latent space modeling framework to adjust for unmeasured confounders, and through a simulation study we show that the LSM performs poorly in terms of Type 1 error when the unmeasured network structure confounds the relationship between the observed covariate and the network.

@@

 Most of these developments have represented new forms for the latent space and/or the further development of the LSM as a Bayesian hierarchical model. In the next section we propose an approach to specifying the Bayesian prior in order to improve inference regarding covariate parameters.
